MG7: Configurable and scalable 16S metagenomics data analysis -- new methods optimized for massive cloud computing

### 2015-09-12 docs/paper/out/paper.tex
generate pdf, refs working -- paper

Metagenomics data analysis is growing at exponential rate during the
last years. The increasing throughput of massively parallel sequencing
technologies, the derived decreasing cost, and the high impact of
metagenomics studies, especially in human health (diagnostics,
treatments, drug response, prevention), are crucial reasons responsible
for this growth of Metagenomics. There is a growing interest in
sequencing all kind of microbiomes (gut, mouth, skin, urinary tract,
airway, milk, bladder), in different conditions of health and disease,
or after different treatments. Metagenomics is also impacting
environmental sciences, crop sciences, agrifood sector and biotechnology
in general. This new possibilities for exploring the diversity of
micro-organisms in the most diverse environments is opening many new
research areas but, due to this wide interest, it is expected that the
amount of data will be overwhelming in the short time
{[}@Stephens-2015{]}.

Genome researchers have raised the alarm over big data in the past
\href{http://www.nature.com/news/genome-researchers-raise-alarm-over-big-data-1.17912}{nature
news add ref} but even a more serious challenge might be faced with the
metagenomics boom/ upswing. If we compare metagenomics data with other
genomics data used in clinical genotyping we find a differential
feature: the key role of time. Thus, for example, in some longitudinal
studies, serial sampling of the same patient along several weeks (or
years) is being used for the follow up of some intestinal pathologies,
for studying the evolution of gut microbiome after antibiotic treatment,
or for colon cancer early detection {[}@Zeller-2014{]}. This need of
sampling across time adds more complexity to metagenomics data storage
and demands adapted algorithms to detect state variations across time as
well as idiosyncratic commonalities of the microbiome of each individual
{[}@Franzosa-2015{]}. In addition to the intra-individual sampling-time
dependence, metagenomic clinical test results vary depending on the
specific region of extraction of the clinical specimen. This local
variability adds complexity to the analysis since different
localizations (different tissues, different anatomical regions, healthy
or tumour tissues) are required to have a sufficiently complete
landscape of the human microbiome. Moreover, reanalysis of old samples
using new tools and better reference databases might be also demanded
from time to time.

During the last years other sciences as astronomy or particle physics
are facing the big data challenge but, at least, these science have
standards for data processing {[}@Stephens-2015{]}. Global standards for
converting raw sequence data into processed data are not yet well
defined in metagenomics and there are shortcomings derived from the fact
that many bioinformatics methodologies currently used for metagenomics
data analysis were designed for a scenario very different that the
current one. These are some of the aspects that have suffered crucial
changes and advances with a direct impact in metagenomics data analysis.
i. The first aspect is related to the sequences to be analyzed: the
reads are larger, the sequencing depth and the number of samples of each
project are considerably bigger. The first metagenomics studies were
very local projects, while nowadays the most fruitful studies are done
at a global level (international, continental, national). This kind of
global studies has yielded the discovery of clinical biomarkers for
diseases of the importance of cancer, obesity or inflammatory bowel
diseases and has allowed exploring the biodiversity in many earth
environments ii. The second aspect derives from the impressive genomics
explosion, its effect being felt in this case in the reference
sequences. The immense amount of sequences available in public
repositories demands new approaches in curation, update and storage for
metagenomics reference databases: current models will or already have
problems to face the future avalanche of metagenomic sequences. iii. The
third aspect to consider for metagenomics data analysis is related to
the appearance of new models for massive computation and storage and to
the new programming methodologies (Scala, \ldots{}) and new cloud models
and resources. The immense new possibilities that these advances offer
must have a direct impact in the metagenomics data analysis. iv. And
finally the new social manner to do science, and especially genomic
science is the fourth aspect to consider. Metagenomics evolves in a
social and global scenario following a science democratization trend in
which many small research groups from distant countries share a common
big metagenomics project. This global cooperation demands systems
allowing following exactly the same pipelines using equivalent cloud
resources to modularly execute the analysis in an asynchronous way of
working between different groups. This definitively new scenario demands
new methods and tools to handle the current and future volume of
metagenomic data with the sufficient speed of analysis. Considering all
these aspects we have designed a new open source methodology for
analyzing metagenomics data that exploits the new possibilities that
cloud computing offers to get a system robust, programmatically
configurable, modular, distributed, flexible, scalable and traceable in
which the biological databases of reference sequences can be easily
updated and/or frequently substituted by new ones or by databases
specifically designed for focused projects.

@@

Our 16S Reference Database is a curated subset of sequences from NCBI
nucleotide database \textbf{nt}. This subset of 16S sequences was
selected by similarity with the bacterial and archaeal reference
sequences downloaded from RDP database {[}@Cole-2014{]}. RDP unaligned
sequences were used to capture new16S sequences from nt using BLAST
similarity strategies and, then, performing additional curation steps to
remove sequences with poor taxonomic assignments to taxonomic nodes
close to the root of the taxonomic tree. All the nucleotide sequences
included in nt database has a taxonomic assignment provided by the
genbank sequence submitter. NCBI provides a table (available at
ftp://ftp.ncbi.nlm.nih.gov/pub/taxonomy/) to do the mapping of any
Genbank Identifier (GI) to its Taxonomy Identifier (TaxID). Thus, we are
based on a submitter-maintained taxonomic annotation system for
reference sequences that supposes a sustainable system able to face the
expected number of reference sequences that will populate the public
global nucleotide databases in the near future. Another advantageous
point is that we are based on NCBI taxonomy, the de facto standard
taxonomic classification for biomolecular data {[}@Cochrane-2010{]}.
NCBI taxonomy is, without any doubt, the most used taxonomy all over the
world and the most similar to the official taxonomies of each specific
field. This is a crucial point because all the type-culture and tissue
databanks follow this official taxonomical classification and, in
addition, all the knowledge accumulated is referred to this taxonomy. In
addition NCBI provides a direct connection between taxonomical formal
names and the physical specimens that serve as exemplars for the species
{[}@Federhen-2015{]}.

@@

\section{7 Tables and Figures}\label{tables-and-figures}

\bibliographystyle{frontiersinSCNS_ENG_HUMS} % for Science, Engineering and Humanities and Social Sciences articles, for Humanities and Social Sciences articles please include page numbers in the in-text citations
%\bibliographystyle{frontiersinHLTH&FPHY} % for Health and Physics articles
\bibliography{refs.bib}

>>>>

Metagenomics data analysis is growing at exponential rate during the
last years. The increasing throughput of massively parallel sequencing
technologies, the derived decreasing cost, and the high impact of
metagenomics studies, especially in human health (diagnostics,
treatments, drug response, prevention), are crucial reasons responsible
for this growth of Metagenomics. There is a growing interest in
sequencing all kind of microbiomes (gut, mouth, skin, urinary tract,
airway, milk, bladder), in different conditions of health and disease,
or after different treatments. Metagenomics is also impacting
environmental sciences, crop sciences, agrifood sector and biotechnology
in general. This new possibilities for exploring the diversity of
micro-organisms in the most diverse environments is opening many new
research areas but, due to this wide interest, it is expected that the
amount of data will be overwhelming in the short time
\citep{Stephens-2015}.

Genome researchers have raised the alarm over big data in the past
\href{http://www.nature.com/news/genome-researchers-raise-alarm-over-big-data-1.17912}{nature
news add ref} but even a more serious challenge might be faced with the
metagenomics boom/ upswing. If we compare metagenomics data with other
genomics data used in clinical genotyping we find a differential
feature: the key role of time. Thus, for example, in some longitudinal
studies, serial sampling of the same patient along several weeks (or
years) is being used for the follow up of some intestinal pathologies,
for studying the evolution of gut microbiome after antibiotic treatment,
or for colon cancer early detection \citep{zeller2014potential}. This
need of sampling across time adds more complexity to metagenomics data
storage and demands adapted algorithms to detect state variations across
time as well as idiosyncratic commonalities of the microbiome of each
individual \citep{Franzosa-2015}. In addition to the intra-individual
sampling-time dependence, metagenomic clinical test results vary
depending on the specific region of extraction of the clinical specimen.
This local variability adds complexity to the analysis since different
localizations (different tissues, different anatomical regions, healthy
or tumour tissues) are required to have a sufficiently complete
landscape of the human microbiome. Moreover, reanalysis of old samples
using new tools and better reference databases might be also demanded
from time to time.

During the last years other sciences as astronomy or particle physics
are facing the big data challenge but, at least, these science have
standards for data processing \citep{Stephens-2015}. Global standards
for converting raw sequence data into processed data are not yet well
defined in metagenomics and there are shortcomings derived from the fact
that many bioinformatics methodologies currently used for metagenomics
data analysis were designed for a scenario very different that the
current one. These are some of the aspects that have suffered crucial
changes and advances with a direct impact in metagenomics data analysis.
i. The first aspect is related to the sequences to be analyzed: the
reads are larger, the sequencing depth and the number of samples of each
project are considerably bigger. The first metagenomics studies were
very local projects, while nowadays the most fruitful studies are done
at a global level (international, continental, national). This kind of
global studies has yielded the discovery of clinical biomarkers for
diseases of the importance of cancer, obesity or inflammatory bowel
diseases and has allowed exploring the biodiversity in many earth
environments ii. The second aspect derives from the impressive genomics
explosion, its effect being felt in this case in the reference
sequences. The immense amount of sequences available in public
repositories demands new approaches in curation, update and storage for
metagenomics reference databases: current models will or already have
problems to face the future avalanche of metagenomic sequences. iii. The
third aspect to consider for metagenomics data analysis is related to
the appearance of new models for massive computation and storage and to
the new programming methodologies (Scala, \ldots{}) and new cloud models
and resources. The immense new possibilities that these advances offer
must have a direct impact in the metagenomics data analysis. iv. And
finally the new social manner to do science, and especially genomic
science is the fourth aspect to consider. Metagenomics evolves in a
social and global scenario following a science democratization trend in
which many small research groups from distant countries share a common
big metagenomics project. This global cooperation demands systems
allowing following exactly the same pipelines using equivalent cloud
resources to modularly execute the analysis in an asynchronous way of
working between different groups. This definitively new scenario demands
new methods and tools to handle the current and future volume of
metagenomic data with the sufficient speed of analysis. Considering all
these aspects we have designed a new open source methodology for
analyzing metagenomics data that exploits the new possibilities that
cloud computing offers to get a system robust, programmatically
configurable, modular, distributed, flexible, scalable and traceable in
which the biological databases of reference sequences can be easily
updated and/or frequently substituted by new ones or by databases
specifically designed for focused projects.

@@

Our 16S Reference Database is a curated subset of sequences from NCBI
nucleotide database \textbf{nt}. This subset of 16S sequences was
selected by similarity with the bacterial and archaeal reference
sequences downloaded from RDP database \citep{Cole-2014}. RDP unaligned
sequences were used to capture new16S sequences from nt using BLAST
similarity strategies and, then, performing additional curation steps to
remove sequences with poor taxonomic assignments to taxonomic nodes
close to the root of the taxonomic tree. All the nucleotide sequences
included in nt database has a taxonomic assignment provided by the
genbank sequence submitter. NCBI provides a table (available at
ftp://ftp.ncbi.nlm.nih.gov/pub/taxonomy/) to do the mapping of any
Genbank Identifier (GI) to its Taxonomy Identifier (TaxID). Thus, we are
based on a submitter-maintained taxonomic annotation system for
reference sequences that supposes a sustainable system able to face the
expected number of reference sequences that will populate the public
global nucleotide databases in the near future. Another advantageous
point is that we are based on NCBI taxonomy, the de facto standard
taxonomic classification for biomolecular data \citep{Cochrane-2010}.
NCBI taxonomy is, without any doubt, the most used taxonomy all over the
world and the most similar to the official taxonomies of each specific
field. This is a crucial point because all the type-culture and tissue
databanks follow this official taxonomical classification and, in
addition, all the knowledge accumulated is referred to this taxonomy. In
addition NCBI provides a direct connection between taxonomical formal
names and the physical specimens that serve as exemplars for the species
\citep{Federhen-2015}.

@@

\bibliographystyle{frontiersinSCNS_ENG_HUMS} % for Science, Engineering and Humanities and Social Sciences articles, for Humanities and Social Sciences articles please include page numbers in the in-text citations
%\bibliographystyle{frontiersinHLTH&FPHY} % for Health and Physics articles
\bibliography{refs}
